RNeXML: a package for reading and writing richly annotated phylogenetic, character, and trait data in R

### 2015-01-07 manuscripts/manuscript.Rmd
Rewrite of the abstract.
This is a little longer than before, but the MEE author guidelines state
350 words as limit, and we remain under that.

---
#layout: preprint
layout: review, 11pt
title: "RNeXML: a package for reading and writing richly annotated phylogenetic, character, and trait data in R"
author: 
  - name: Carl Boettiger
    affiliation: cstar
    email: cboettig(at)gmail.com
    footnote: Corresponding author
  - name: Hilmar Lapp
    affiliation: NESCent
  - name: Scott Chamberlain
    affiliation: ropensci
  - name: Rutger Vos
    affiliation: NBC
address: 
  - code: cstar
    address: | 
      Center for Stock Assessment Research, 
      Department of Applied Math and Statistics, 
      University of California, Mail Stop SOE-2,
      Santa Cruz, CA 95064, USA
  - code: NESCent
    address: | 
      National Evolutionary Synthesis Center, Durham, NC, USA
  - code: ropensci
    address: |
      University of California, Berkeley, CA, USA
  - code: NBC
    address: | 
      Naturalis Biodiversity Center, Leiden, the Netherlands
abstract: | 
      1. NeXML is a new exchange format for phylogenetic data and
      metadata that offers substantial improvements in reliability,
      extensibility, and richness over earlier formats such as
      NEXUS. In this paper we present the RNeXML package which
      provides a bridge between this data format and the extensive
      library of phylogenetic tools already available in the popular
      data analysis environment and programming language R, with a
      particular emphasis on comparative methods.

      2. The ability to parse and write NeXML from within R is only a
      small part of the functionality offered by the package.  We
      illustrate how RNeXML facilitates adding and manipulating data
      annotations whose meaning can be defined in a precise and
      machine-interpretable way by means of formal ontologies and other
      semantic web technologies.

      3. An essential feature of NeXML is to provide both the
      flexibility to be continually extended to meet the emerging
      needs of the phylogenetics community without breaking software
      unaware of such extensions. We illustrate how RNeXML makes
      it easy to extend NeXML in this manner.

      4. RNeXML is well suited for users to preserve and maintain
      their personal archives of phylogenetic and comparative trait
      data in consistent, query-able, and cross-platform format, to
      leverage the rapidly growing public community archives of rich
      NeXML files, and to interact with web services that return data
      and their metadata packaged together as NeXML.

@@

This project was supported in part by the National Evolutionary
Synthesis Center (NESCent) (NSF #EF-0905606), and grants from the
Natioal Science Foundation (DBI-1306697) and the Alfred P Sloan
Foundation (Grant 2013-6-22). `RNeXML` started as a project idea for
the Google Summer of Code, and we thank Kseniia Shumelchyk for taking
the first steps to implement it. We are grateful to F. Michonneau for
helpful comments on an earlier version of this manuscript.

@@

<!-- Footnotes -->

[^1]: By providing the optional `dependencies` argument, R will install
additional packages used in some of the advanced examples shown later
but which are not required for basic functionality.

[^2]: For convenience `nexml_read` is also aliased as `read.nexml`,
for users familiar with the somewhat analogous `read.nexus` function.

>>>>

---
#layout: preprint
layout: review, 11pt
title: "RNeXML: a package for reading and writing richly annotated phylogenetic, character, and trait data in R"
author: 
  - name: Carl Boettiger
    affiliation: cstar
    email: cboettig(at)gmail.com
    footnote: Corresponding author
  - name: Hilmar Lapp
    affiliation: NESCent
  - name: Scott Chamberlain
    affiliation: ropensci
  - name: Rutger Vos
    affiliation: NBC
address: 
  - code: cstar
    address: | 
      Center for Stock Assessment Research, 
      Department of Applied Math and Statistics, 
      University of California, Mail Stop SOE-2,
      Santa Cruz, CA 95064, USA
  - code: NESCent
    address: | 
      National Evolutionary Synthesis Center, Durham, NC, USA
  - code: ropensci
    address: |
      University of California, Berkeley, CA, USA
  - code: NBC
    address: | 
      Naturalis Biodiversity Center, Leiden, the Netherlands
abstract: | 
      1. NeXML is a powerful and forward-compatible exchange standard
      recently proposed to better meet the expanding needs for
      machine-proof phylogenetic data and metadata sharing. Here we
      present the RNeXML package, which aims to provide to users of
      the popular R programming language easy programmatic access to
      reading and writing NeXML documents, including rich metadata, in
      a way that interfaces seamlessly with the extensive library of
      phylogenetic tools already available in the R ecosystem.

      2. Wherever possible, we designed RNeXML to map NeXML document
      contents, whose arrangement is influenced by the format's
      normative XML Schema definition, to their most intuitive or
      useful representation in R. To make NeXML's powerful facility
      for recording semantically rich machine-readable metadata
      accessible to R users, we designed a functional programming
      interface to it that hides the semantic web standards leveraged
      by NeXML from R users who are unfamiliar with them.

      3. RNeXML can read any NeXML document that validates, and it
      generates valid NeXML documentes from phylogeny and character
      data from various R representations in use. The metadata
      programming interface at a basic level aids fulfilling data
      documentation best practices, and at an advanced level preserves
      NeXML's nearly limitless extensibility, for which we provide a
      fully working demonstration. Furthermore, to lower the barriers
      to sharing well-documented phylogenetic data, RNeXML has started
      to integrate with taxonomic metadata augmentation services on
      the web, and with online repositories for data archiving.

      4. RNeXML allows R's rich ecosystem to read and write data in
      the NeXML format through an interface that is no more involved
      than reading or writing data from other, less powerful data
      formats. It also provides an interface designed to feel familiar
      to R programmers and to be consistent with recommended practices
      for R package development, yet that retains the full power for
      users to add their own custom data and metadata to the
      phylogenies they work with, without introducing potentially
      incompatible changes to the exchange standard.

@@

This project was supported in part by the National Evolutionary
Synthesis Center (NESCent) (NSF #EF-0905606), and grants from the
Natioal Science Foundation (DBI-1306697) and the Alfred P Sloan
Foundation (Grant 2013-6-22). `RNeXML` started as a project idea for
the Google Summer of Code(TM), and we thank Kseniia Shumelchyk for taking
the first steps to implement it. We are grateful to F. Michonneau for
helpful comments on an earlier version of this manuscript.

@@
